\begin{vitae}
\addcontentsline{toc}{chapter}{Curriculum Vitae}
\vspace{-0.4cm}

\begin{vitaesection}{Education}
\vspace{-0.1cm}
\item [2023] 	Ph.D. in Computer Science, University of California, Santa Barbara
\item [2022]	M.S. in Computer Science, University of California, Santa Barbara
\item [2016]	B.A. in Computer Science, University of Virginia
\item [2016]	B.A. in Physics, University of Virginia
\end{vitaesection}

\textbf{Publications}
\begin{enumerate}
    \item Isaac Mackey and Jianwen Su. 2023. Early detection of temporal constraint violations (invited extension, in revision). Information and Computation, Special issue: ``Temporal Representation and Reasoning,'' (2023).
    \item Isaac Mackey and Jianwen Su. 2023. Mapping singly-linked rules to linear temporal logic formulas. Information Systems (2023), 102222.
    \item Isaac Mackey, Raghubir Chimni, and Jianwen Su. 2022. Early detection of temporal constraint violations. In 29th International Symposium on Temporal Representation and Reasoning (TIME 2022). Schloss Dagstuhl-Leibniz-Zentrum fur Informatik.
    \item Isaac Mackey and Jianwen Su. 2022. A Rule-Based Constraint Language for Event Streams. In 4th International Workshop on the Resurgence of Datalog in Academia and Industry. 145-150.
    \item Moriba Jah, Emine Basak Alper Ramaswamy, Jeffrey Stuart, Isaac Mackey, and James Crowley. 2020. Improving Navigation Analysis with OD-D: The Visually Interactive Orbit Determination Dashboard. Jet Propulsion Laboratory, NASA, Pasadena, California. (2020).
    \item Isaac Mackey and Jianwen Su. 2019. Mapping Business Rules to LTL Formulas. In: Service-Oriented Computing: 17th International Conference, ICSOC 2019, Toulouse, France, October 28-31, 2019, Proceedings. Springer Lecture Notes in Computer Science, 563-565.
    \item Yan Tang, Isaac Mackey, and Jianwen Su. 2018. Querying workflow logs. Information 9, 2 (2018), 25.
    \item Ernst L. Leiss and Isaac Mackey. 2018. The worst parallel Hanoi graphs. Theoretical Computer Science, Volume 705 (2018), 1-8.
\end{enumerate}

\textbf{Professional Experience}
\begin{itemize}
    \item Commissioned as a Second Lieutenant in the United States Marine Corps \hfill{2023}

    \item Internship in software development, Naval Information Warfare Center Pacific \hfill{2021}
    % Software development for scheduling tool to maintain mission-capable aircraft.
    % Front-end design with JavaScript, React, Chakra. Agile methodology.
    
    \item Internship in software development, NASA Jet Propulsion Laboratory \hfill{2019}
    % Software development for orbit determination tool
    % to aid mission planning and spacecraft trajectory correction.
    % Front-end design with JavaScript, d3. Back-end with Node.js.
    % Meetings with Mission Design, Navigation experts.
\end{itemize}

\end{vitae}